---
title: "Econ294A R programming final"
author: "Feng Hong"
date: "March 17, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
library(Rcpp)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(class)
library(nycflights13)  #for data
library("RSQLite") #for sqllite
nycflights13_sqlite()
```

## Extract Data from Database
```{r raw, message=FALSE}
flights_sqlite <- tbl(nycflights13_sqlite(), "flights")
airlines_sqlite <- tbl(nycflights13_sqlite(), "airlines")
airports_sqlite <- tbl(nycflights13_sqlite(), "airports")
planes_sqlite <- tbl(nycflights13_sqlite(), "planes")
weather_sqlite <- tbl(nycflights13_sqlite(), "weather")
```

## Join the table *flights* and *planes*
```{r flights_planes_join, message=FALSE}
inner_flights_planes <- inner_join(flights, planes, by = "tailnum") %>% tbl_df 
names(inner_flights_planes)
colnames(inner_flights_planes)[1] <- "flight_year"
colnames(inner_flights_planes)[15] <- "dep_hour"
colnames(inner_flights_planes)[17] <- "plane_year"
```

## Create the date index
```{r data_index_creation}
flights_planes <- inner_flights_planes %>% 
  mutate(
    date = paste(flight_year, month, day, sep = "-"), 
    date = as.Date(date, format = "%Y-%m-%d"), # create date to merge with weather
    cancelled = ifelse(is.na(arr_time), 1, 0)  # question requires this
  ) 
```

## Change character variable to factor variable
```{r select_columns}
flights_planes$carrier <- as.factor(flights_planes$carrier)
flights_planes$origin <- as.factor(flights_planes$origin)
flights_planes$dest <- as.factor(flights_planes$dest)
flights_planes$manufacturer <- as.factor(flights_planes$manufacturer)
```

## Change integer variable to factor variable
``` {r type_conversion}
flights_planes$month <- as.factor(flights_planes$month)
flights_planes$flight <- as.factor(flights_planes$flight)
```


## Weather Data Refinement
```{r weather1}
weather <- weather_sqlite %>%  
  collect() %>% 
  mutate(
    date = paste(year, month, day, sep = "-"),
    date = as.Date(date, format = "%Y-%m-%d"),
    weekday = weekdays(date),
    weekday = as.factor(weekday) # add the weekday variable
  ) 
```


```{r weather2}
weather_mean <- weather %>% group_by(date) %>%
  summarise(
    weekday = first(weekday),
    mean_temp = mean(temp),
    mean_dewp = mean(dewp),
    mean_humid = mean(humid),
    # mean_wind_dir = mean(wind_dir), wind direction has too many NA's.
    mean_wind_speed = mean(wind_speed),
    mean_wind_gust = mean(wind_gust),
    mean_precip = mean(precip),
    # mean_pressure = mean(pressure), pressure has too many NA's.
    mean_visib = mean(visib)
  )
```


## Identify the highly correlated data
```{r correlated_data}
corr_weather_mean <- cor(na.omit(weather_mean[,3:9]))
print(corr_weather_mean)
```

## Delete columns that are highly correlated from weather_mean
```{r data_delete}
weather2 <- weather_mean %>%
  dplyr::select(date, weekday, mean_temp, mean_wind_speed, mean_precip, mean_visib)
corr_weather2<- cor(na.omit(weather2[,3:6]))
print(corr_weather2)
```

## Join flights, planes and weather data and obtain the final data
```{r final_data}
final_data <- inner_join(flights_planes, weather2, by = "date")
length(final_data[final_data$cancelled == 1])
names(final_data)
```


## **OLS** for *dep_delay*
```
I expect the following regressors to be significant. Some similar variables 
such as mean_temp, mean_wind_speed are not included as arguments mostly because
their impact on departure delay can be ambiguous. 
```

```{r OLS}
model.delay <- lm(dep_delay ~ month + as.factor(weekday) + carrier + origin +
                plane_year + seats + mean_precip + mean_visib,
               data = final_data)
summary(model.delay)
```

```
From the regression we can see that dummy variables "month" and "carrier" are
generally significant regressor. "Weekday, "origin, "plane_year", "seats",
"mean_precip", "mean_visib" are all significant. The coefficients stated in the
regression results.
However, the model has an Adjusted R-squared as small as 0.05242. So based on
this model, we might not be able to predict the departure delay precisely.
```


## **GLS** for *cancel*
```
For regression of cancellation, use logit model because "cancelled" is a binary
variable.
```

```{r GLS}
model.cancel <- glm(cancelled ~ month + as.factor(weekday) + carrier + origin +
                     seats +  mean_visib,
                   data = final_data, family=binomial(link="logit"))
summary(model.cancel)
```

```
I’ve tried several other regressors, which all turned out to be insignificant. 
In this particular model above, the factors chosen are almost all significant,
the coefficients stated in the regression results.
Using this model, we can predict the probability of cancellation based on the
independent variabls.
```

```
I'm also interested in how different weather conditions would affect the daily
mean cancellation. So I find daily total cancellation and keep weather variables
to run the regression.  
```

```{r time, Warning=FALSE}
total_cancel_by_date <- final_data %>% group_by(date) %>% 
  summarise(cancel_total = sum(cancelled),
            mean_temp = mean(mean_temp, na.rm = T),
            mean_wind_speed = mean(mean_wind_speed, na.rm = T),
            mean_precip = mean(mean_precip, na.rm = T),
            mean_visib = mean(mean_visib, na.rm = T)) 

model.total_cancel <- lm(cancel_total ~ mean_temp + mean_wind_speed + 
                           mean_precip + mean_visib, 
                         data = total_cancel_by_date)
summary(model.total_cancel)

```

```
The independent variable mean_wind_speed and mean_visib are both significant.
Using this model we can predict how many cancellations might occur based on
weather conditions. But again, adjusted r-squared is not high enough to justify
the model.
```

## Plotting with ggplot2
# (a) weather

```{r plot_a}
plot_weather <- ggplot(data = final_data, aes(mean_precip, mean_visib)) 
plot_weather + geom_point( aes(color = as.factor(cancelled)),size = 1) +
  xlab("Precipitation") + ylab("Visibility")
```

```
The plot shows a simple relationship between cancellation and 2 major weather
conditions: visibility and precipitation. As we can see for most sunny clear
days, cancellation relatively rarely happens.
```

# (b) day of week and time of year
```{r plot_b}
month_weekday <- final_data %>%
  group_by(month, weekday) %>%
  summarise(mean.dep_delay = mean(dep_delay, na.rm = T)) 

plot_time <- ggplot(month_weekday, aes(x = month, y= mean.dep_delay))
plot_time + geom_point(aes(color = weekday), size = 3) +
  xlab("Month") + ylab("Departure Delay")
```

```
This picture gives us an idea of the pattern of departure delay with respect to
weekday for each month. Weekend flights seem to have less departure delay.
```
# (c) airport destination

```{r plot_c}
plot_destination <- ggplot(data = final_data, aes(month, dep_delay, color = distance)) 
plot_destination + geom_point() +
  xlab("Month") + ylab("Departure Delay")
```


# (d) characteristics of the plane
```{r plot_d}
plot_plane <- ggplot(data = final_data, aes(x = seats, y = dep_delay, 
                                            color = engines, shape = type))
plot_plane + geom_point(alpha=0.2) +
  xlab("Month") + ylab("Departure Delay")
```

```
The last picture here shows the relationship between engine number, plane type
and departure delay for each month.
```


## Try some data mining - Based on KNN
```
Got the error message:
Error in knn(train = knn_train, test = knn_test, cl = knn_train_target,  : 
'train' and 'class' have different lengths

Have to give up due to limit of time     
KNN reference: https://www.youtube.com/watch?v=DkLNb0CXw84
```
```{r KNN, eval=FALSE}
table(final_data$cancelled)
set.seed(9850)
gp <- runif(nrow(final_data))

final_data_knn <- final_data[order(gp),]

# DEBUG:
# str(final_data_knn)
# summary(final_data_knn)

knn_no_na <- final_data_knn[(!is.na(final_data_knn$mean_temp)) & 
                    (!is.na(final_data_knn$mean_wind_speed)) & 
                    (!is.na(final_data_knn$mean_precip)) & 
                    (!is.na(final_data_knn$mean_visib)),]

normalize <- function(x){ return( (x- min(x))/(max(x) - min(x)) ) }
knn_normal <- as.data.frame(lapply(knn_no_na[, c(28,29,30,31)], normalize))
summary(knn_normal)


knn_train <- knn_normal[1:250000,]
knn_test <- knn_normal[250001:281860,]

# DEBUG:
# names(knn_no_na)
knn_train_target <- knn_no_na[1:250000, 26]
knn_test_target <-knn_no_na[250001:281860, 26]

k = 500
m1 <- knn(train = knn_train, test = knn_test, cl=knn_train_target, k =500)

```
